---
title: "subsetting-short"
author: "Julien Brun"
date: "7/17/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Subsetting in R

For a full overview on the topic, we recommend you consult the [Subsetting chapter of R Advanced](http://adv-r.had.co.nz/Subsetting.html) from H. Wickham. This document is just a quick summary of it to fit in a 30minintro session for NCEAS SciComp interns. For another quick overview, feel free to check: https://www.statmethods.net/management/subset.html

<br>

There are 3 subsetting operators in R: `[`, `[[` and `$`

**=> They behave differently based on how which data type they are operating on!**

In this short session, we will be only focusing on `atomic vectors`, `lists` and `data frames`

<br>

The big difference between `[` and `[[` is that `[` tends to *preserve* what it is subsetting and ``[[` tends to *simplify* it. Note this statement is not fully accurate accross all the data types. 
A good analogy often used is the pepperpot one: With `[` you get the pot, with `[[` you get the pepper inside the pot

<br>

`$` is a useful shorthand for `[[` for character/named subsetting. There’s one important difference between $ and [[. $ does partial matching:

```{r partial matching, eval=FALSE}
x <- list(abc = 1)
x$a
```


## Subsetting Vector

```{r subsetting vectors, eval=FALSE}

# Define a vector
(v <- c("a", "b", "c"))
# [1] 1 2 3

# Subset with []
v[1]
str(v[1])
v[c(1,3)]
v[-2] # drop the position 

# Subset with [[]]
v[[1]]
str(v[[1]])
# v[[c(1,3)]]  # won't work! [[ can only return one element!

identical(v[1], v[[1]]) 
#
```
*=> identical but wait: not true when you add attributes to an atomic vector*
 
```{r subsetting named vectors, eval=FALSE}
# Let us use a named vector to demonstrate this
(v <- c(a = 1, b = 2, c = 3))

# Check the attributes
attributes(v)

# Subset with []
v[1]

# Subset with [[]]
v[[1]]

identical(v[1], v[[1]]) # not anymore identical!
```

## Subsetting List

```{r subsetting lists, eval=FALSE}

# Even more explicit if you used a named vector
(l <- list(a = c(1, 2), b = 2, c = 3))

# Subset by position using []
l[1]
str(l[1]) # you get a list back (preserved)

# Subset by name
l["b"]

# Subset by position using [[]]
l[[2]]
str(l[[2]]) # you get a number back (simplified)

# Subset a specific element of the vector
l[[1]][2]

# Subset by name
l[["b"]]
l$b  # equivalent

# Subset a specific element of the vector
l$a[2]

# Reassign values
l[["b"]] <- c(1,2,3)
l
```

## Subsetting Data Frame

```{r subsetting data frames, eval=FALSE}

# Even more explicit if you used a named vector
(df <- data.frame(a = c(1,2,3), ab = c("a", "ab", "c"), c = c(4, 5, 6), stringsAsFactors = FALSE))

# Subset by position
df[1]
str(df[1])

df[[2]]
str(df[[2]])

# Subset by name
df["ab"]
df[["ab"]]
df$b

# Subset several columns (à la list)
df[c("a", "ab")]

# Subset several columns (à la matrix)
df[ , c("a", "ab")]  

# Note there is an important difference between the two methods if you only subset one column 
df["a"]   # The list way does not simplifies
df[, "a"] # The matrix way does simplifies


# Subset a "cell"
df[1, 1]
# or
df[1, "a"]
df$a[1]


# Selecting columns
myvars <- names(df) %in% c("a", "c")
df[myvars]
df[!myvars]

# Selecting rows
# Using which, i.e. position
which(df$c > 4)
df[which(df$c > 4), ]

# Using condition, i.e. TRUE/FALSE
df$c > 4
df[df$c > 4, ]

# Using a pattern: select coloumns starting with a
df[ , grepl(pattern = "^a", names(df))]

```










